---
title: PMTiles - self created
output: 
  bookdown::html_vignette2:
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
#load_all("../tmap");load_all("../tmap.mapgl");load_all();tmap_mode("maplibre")
```



## Introduction

Until version 4.3, it was only possible to visualize spatial data stored in memory with `tmap`, e.g. objects from the class packages `sf`, `terra`, or `stars`. This package, `tmap.sources` makes it possible to visualize remote spatial data. Currently only `PMTiles`, a remote tile-based data sources are supported, but support for other remote data sources will be added later.

To run `tmap.sources`, besides `tmap`, another package is required, namely [`tmap.mapgl`](https://r-tmap.github.io/tmap.mapgl/articles/mapgl), because we require the tmap mode `"maplibre"`:


```{r}
library(tmap)
library(tmap.mapgl)
library(tmap.sources)
tmap_mode("maplibre")
```

In order to create PMTiles locally, the package `freestiler` is required. The created `PMTiles` files can be viewed from a local server, or be uploaded to a remote server.


```{r}
library(freestiler)
```

## Creation of PMTiles

If we have a `sf` object for which we would like to create a PMTiles file, we have to think in advance what we would like to visualize. To replicate the choropleth in the introduction vignette, we have two options:

1. We generate a color column, and use the as-is scale (`tm_scale_asis`)
2. We generate a categorical (factor) column and use the categorical scale (`tm_scale_categorical`)

Let's do both:

```r
# add categorical variable
NLD_dist$edu_cat = cut(NLD_dist$edu_appl_sci, breaks = c(0, 21.5, 29.5, 39.8, 51.5, 90))

# get used colors
library(cols4all)
pal = c4a("plasma", 5)
colorNA = c4a_na("plasma")

# add color variable
NLD_dist$edu_fill = pal[cut(NLD_dist$edu_appl_sci, breaks = c(0, 21.5, 29.5, 39.8, 51.5, 90), include.lowest = TRUE)]
NLD_dist$edu_fill[is.na(NLD_dist$edu_fill)] = colorNA
```

Now we can create the PMTiles file:

```r
freestile(
	NLD_dist,
	"NLD_dist.pmtiles",
	layer_name = "edu",
	min_zoom = 4,
	max_zoom = 15
)
```

### Option 1: color column

```r
tmap_mode("maplibre")

tm_shape("NLD_dist.pmtiles") +
  tm_polygons(fill = "edu_fill") +
  tm_add_legend(fill = pal, labels = levels(NLD_dist$edu_cat), type = "polygons", title = "Education")
```


### Option 2: categorical variable

```r
tmap_mode("maplibre")

tm_shape("NLD_dist.pmtiles") +
  tm_polygons(
    fill = "edu_cat",
    fill.scale = tm_scale_categorical(values = pal, levels = levels(NLD_dist$edu_cat)),
    fill.legend = tm_legend(title = "Education")
  )
```
