---
title: Self-created PMTiles
output: 
  bookdown::html_vignette2:
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 8,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {
     if (length(x) > lines) {
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

## Introduction

PMTiles is an efficient format for serving vector and raster map tiles directly
from a single file, without a tile server. This vignette shows how to create
PMTiles from `sf` objects using the `freestiler` package, and how to visualize
them with `tmap.mapgl`.

```{r}
library(tmap)
library(tmap.mapgl)
library(tmap.sources)
library(freestiler)
library(cols4all)
tmap_mode("maplibre")
```

Two things to decide before writing a PMTiles file:

- **Which columns to include.** Only columns present in the file can be used for
  styling. Add any variables you intend to map before calling `freestile()`.
- **How to encode the aesthetic.** For categorical styling, either pre-compute a
  color column (`tm_scale_asis`) or include a factor column and use
  `tm_scale_categorical`. Both approaches are shown below for polygons.

## Polygons

### Preparing the data

```{r}
data("NLD_dist", package = "tmap")

breaks  <- c(0, 21.5, 29.5, 39.8, 51.5, 90)
pal     <- c4a("plasma", 5)
colorNA <- c4a_na("plasma")

NLD_dist$edu_cat  <- cut(NLD_dist$edu_appl_sci, breaks = breaks)
NLD_dist$edu_fill <- pal[cut(NLD_dist$edu_appl_sci, breaks = breaks,
                             include.lowest = TRUE)]
NLD_dist$edu_fill[is.na(NLD_dist$edu_fill)] <- colorNA
```

### Creation of PMTiles

```{r}
freestile(
  NLD_dist,
  "NLD_dist.pmtiles",
  layer_name = "edu",
  min_zoom = 4,
  max_zoom = 15
)
```

### Visualization

**Option 1:** use the pre-computed color column directly with `tm_scale_asis`.
The legend must be added manually since tmap cannot infer it from raw color
values.

```{r, eval = FALSE}
tm_shape("NLD_dist.pmtiles") +
  tm_polygons(fill = "edu_fill") +
  tm_add_legend(fill = pal, labels = levels(NLD_dist$edu_cat),
                type = "polygons", title = "Education")
```
```{r, echo = FALSE, fig.height = 4.5}
tm_shape(NLD_dist) +
  tm_polygons(fill = "edu_fill") +
  tm_add_legend(fill = pal, labels = levels(NLD_dist$edu_cat),
                type = "polygons", title = "Education")
```

**Option 2:** use the factor column with `tm_scale_categorical`. tmap handles
the color mapping and legend automatically.

```{r, eval = FALSE}
tm_shape("NLD_dist.pmtiles") +
  tm_polygons(
    fill        = "edu_cat",
    fill.scale  = tm_scale_categorical(values = pal,
                                       levels = levels(NLD_dist$edu_cat)),
    fill.legend = tm_legend(title = "Education")
  )
```
```{r, echo = FALSE, fig.height = 4.5}
tm_shape(NLD_dist) +
  tm_polygons(
    fill        = "edu_cat",
    fill.scale  = tm_scale_categorical(values = pal,
                                       levels = levels(NLD_dist$edu_cat)),
    fill.legend = tm_legend(title = "Education")
  )
```

## Lines

### Preparing the data

`World_rivers` contains a `scalerank` variable (1 = largest rivers). We keep
the six most prominent ranks and convert to a factor for categorical styling.

```{r}
data("World_rivers", package = "tmap")
World_rivers <- World_rivers[World_rivers$scalerank <= 6, ]
World_rivers$rank_cat <- factor(paste0("rank ", World_rivers$scalerank),
                                levels = paste0("rank ", 1:6))
river_pal <- c4a("-seaborn.crest", 6)
```

### Creation of PMTiles

```{r}
freestile(World_rivers, "World_rivers.pmtiles",
          layer_name = "rivers", min_zoom = 0, max_zoom = 10)
```

### Visualization

```{r, eval = FALSE}
tm_shape("World_rivers.pmtiles") +
  tm_lines(
    lwd        = 6,
    col        = "rank_cat",
    col.scale  = tm_scale_categorical(values = river_pal,
                                      levels = levels(World_rivers$rank_cat)),
    col.legend = tm_legend(title = "River rank")
  )
```
```{r, echo = FALSE, fig.height = 4.5}
tm_shape(World_rivers) +
  tm_lines(
    lwd        = 6,
    col        = "rank_cat",
    col.scale  = tm_scale_categorical(values = river_pal,
                                      levels = levels(World_rivers$rank_cat)),
    col.legend = tm_legend(title = "River rank")
  )
```

## Points

When creating point PMTiles from polygon data, centroids must be computed
first. Passing polygon geometry to `freestile()` for use with `tm_dots()` or
`tm_symbols()` will result in circles rendered at every vertex rather than at
the feature centroid.

### Preparing the data

```{r}
NLD_dist$edu_cat <- cut(NLD_dist$edu_appl_sci, breaks = breaks)

# centroid step required before freestile() when source is polygon sf
NLD_pts <- sf::st_centroid(NLD_dist)
```

### Creation of PMTiles

```{r}
freestile(NLD_pts, "NLD_dist_pts.pmtiles",
          layer_name = "districts", min_zoom = 5, max_zoom = 15)
```

### Visualization

```{r, eval = FALSE}
tm_shape("NLD_dist_pts.pmtiles") +
  tm_dots(
    fill        = "edu_cat",
    fill.scale  = tm_scale_categorical(values = pal,
                                       levels = levels(NLD_dist$edu_cat)),
    fill.legend = tm_legend(title = "Education")
  )
```
```{r, echo = FALSE, fig.height = 4.5}
tm_shape(NLD_pts) +
  tm_dots(
    fill        = "edu_cat",
    fill.scale  = tm_scale_categorical(values = pal,
                                       levels = levels(NLD_dist$edu_cat)),
    fill.legend = tm_legend(title = "Education")
  )
```

## 3D Polygons

Extruded polygons require a numeric height variable. Because the raw data range
is not available from the remote tile, it must be declared explicitly via
`tm_scale_continuous(limits = ...)`. Use `range()` on your data beforehand to
find the appropriate values.

### Preparing the data

```{r}
NLD_dist$edu_fill <- pal[as.integer(
  cut(NLD_dist$edu_appl_sci, breaks = breaks, include.lowest = TRUE)
)]
NLD_dist$edu_fill[is.na(NLD_dist$edu_fill)] <- colorNA
NLD_dist$pop_density <- NLD_dist$population / as.numeric(NLD_dist$area)
pop_density_range    <- range(NLD_dist$pop_density, na.rm = TRUE)
```

### Creation of PMTiles

The same `NLD_dist.pmtiles` written above already contains `pop_density`, so
no new file is needed here — provided `freestile()` was called after adding the
column.

```{r, eval = FALSE}
# only needed if NLD_dist.pmtiles was created before pop_density was added
freestile(NLD_dist, "NLD_dist.pmtiles",
          layer_name = "edu", min_zoom = 4, max_zoom = 15)
```

### Visualization

```{r, eval = FALSE}
tm_shape("NLD_dist.pmtiles") +
  tm_polygons_3d(
    fill         = "edu_fill",
    height       = "pop_density",
    height.scale = tm_scale_continuous(limits = pop_density_range),
    fill.legend  = tm_legend(title = "Education (% applied sci.)"),
    options      = opt_tm_polygons_3d(height.min = 2000)
  ) +
  tm_maplibre(pitch = 45)
```
```{r, echo = FALSE, fig.height = 4.5}
tm_shape(NLD_dist) +
  tm_polygons_3d(
    fill         = "edu_fill",
    height       = "pop_density",
    height.scale = tm_scale_continuous(limits = pop_density_range),
    fill.legend  = tm_legend(title = "Education (% applied sci.)"),
    options      = opt_tm_polygons_3d(height.min = 2000)
  ) +
  tm_maplibre(pitch = 45)
```
